---
title: "An Introduction to ggplot2 (Workshop)"
author: "Elena"
date: '2022-05-17'
slug: an-introduction-to-ggplot2
categories: R
tags:
- visualisation
- colour
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
library(RColorBrewer)
library(tidyverse)
library(details)
```

This is an adaptation of a workshop I ran last year as part of the Centre of Excellence for the Dynamics of Language (ANU) Seminar Series- the main change is that I've subbed out the data with the iris data that's automatically available in R so that it's more accessible. If you would like a recording of the workshop, which uses vowel data, you can email me at elena.sheard@anu.edu.au  

\foldstart{[## Introduction to ggplot2]}
###   Workshop Overview 
- Introduction to the ggplot2 package and its main functions
- How to make three plots
  + Scatter plot
  + Box plot
  + Density plot
- How to customise these plots 
  + Colour
  + Labels 
  + Legends
  + Themes
  + Faceting
  + Manual palettes
  
### The Basics
- ggplot2 is a R package dedicated to data visualisation
- Has an underlying grammar that allows you to build graphs by combining independent components
  + This allows you to build almost any type of chart
  + The same underlying data can be transformed by many different scales or layers (i.e., it is extremely flexible)
- It is also over a decade old, meaning there are a lot of resources available

- To use ggplot2, you need to:
  + Install and load either the ggplot2 package or the tidyverse package 
  + Load the data (`name_of_object <- read.csv("name_of_spreadsheet.csv")`)
- I have also included code for installing and loading the RColorBrewer package

```{r,echo=TRUE, eval=FALSE}
install.packages("tidyverse") 
install.packages("RColorBrewer") 
library(RColorBrewer)
library(tidyverse)
```

### The base layer

- All plots are composed of:
  + The **data**: the information you want to visualise
  + A **mapping**: a description of how you want the variables in your data to be 'mapped' to aesthetic attributes like colour, shapes, or x and y axes
- All plots you make in ggplot2 will begin with the `ggplot()` function  
  + This builds the first component of your graph (the base layer)  
- You also need to tell ggplot what data you want to visualise 
  + The name of your dataframe or object, in our case 'cars'  
- The code below will create an empty base layer
  
```{r,eval=TRUE}
ggplot(data=iris)

```

### Mapping 
- Mapping depends on what kind of graph you are after, but for *most* you will want to add **x and/or y axes**
- Within `ggplot(data=ban)`, we need another function `aes()` within which you give the x and y axes so that:
  + `ggplot(data=*dataframe*, aes(x=*column_a*, y=*column_b*))` 
  + The names you give for 'x' and 'y' are the **names of the columns in the dataframe you want to plot**  
- The code below will create a graph with a labeled x and y axis that is otherwise empty. In the next section, we will turn it into a scatter plot

```{r, eval=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length))
```
\foldend

## Three Basic Plots 
### Basic scatter plot

- We already have the basis for a scatterplot from the previous code as we have **two continuous variables** on the x and y axes
- To turn this into a scatterplot, add `+ geom_point()` to the previous code
  + This tells ggplot2 that we want to build a scatter plot, with the specified x and y axes
- When you add a new component to a graph, there must be a `+` connecting to the previous one
- And voila! 

```{r,eval=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length))+
  geom_point()
```
  
### Basic boxplot

- The basic `ggplot(data=, aes(x=,y=))` stays the same
  + Instead of `+geom_point()` we add `+geom_boxplot()`
- Box plots use:
  + a **categorical** variable on the **x axis** 
  + a **continuous** variable on the **y axis**

```{r}
ggplot(data=iris, aes(x=Species,y=Petal.Length))+ 
  geom_boxplot()
```

### Basic density plot 

- The basic code is `ggplot(data=, aes(x=))` 
  + This time we add `+geom_density()`
- Density plots use:
  + a **single continuous** variable on the **x-axis**
  + Y axis tells you the distribution of this variables
  
```{r}
ggplot(data=iris, aes(x=Petal.Length))+
  geom_density()
```

## Practice Mini-Tasks {.tabset}
### 1: Build a scatter plot 

- Make your own scatter plot by changing the x and y axes to different continuous variables from the data:
  + Petal.Length
  + Petal.Width 
  + Sepal.Length
  + Sepal.Width
  
### 2: Build a box plot

- Make your own box plot by changing the y axis to different continuous variables from the dataframe:
  + Petal.Length
  + Petal.Width 
  + Sepal.Length
  + Sepal.Width

- And change the x axis to a different categorical variable from the dataframe (although with this data there is only one option)
  + Species
  
### 3: Build a density plot 

- Make your own density plot by changing the x axis to a different continuous variable from the dataframe
  + Petal.Length
  + Petal.Width 
  + Sepal.Length
  + Sepal.Width
  
## Customising in ggplot2 {.tabset}
### Colour 
- Colour is a very easy way to add additional information 
- We map colour to variables within the `aes()` function, after we have put in the x and y axes   
  + `ggplot(data=data_frame, aes(x=columm_a,y=column_b,color=column_c))`
- The same principle applies to shape and linetype 

```{r,}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length, color=Species))+
  geom_point()
```

- If your graph has shapes with lines (like boxplots and density plots), you can:
  + change the line colour with `color=column_c` 
  + change the colour inside the lines with `fill=column_d`
  
```{r, }
ggplot(data=iris, aes(x=Sepal.Length,color=Species))+
  geom_density()
```

```{r, }
ggplot(data=iris, aes(x=Sepal.Length,fill=Species))+
  geom_density()
```
### Labels and Titles

- The `+labs()` function lets you change the plot title, caption, x and y axes and the legend labels
  + `labs(title="Plot Title",caption="Plot Caption",x="column_a",y="column_b", color="column_c",fill="column_d")`
- To change the legend labels, you specify the mapping attribute (color or fill), then the column being mapped to that attribute

```{r,}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length,color=Species))+
  geom_point()+
  labs(title="Sepal Length and Petal Length",
      caption="Data from R", 
      x="Sepal Length",y="Petal Length",
      color="Species")
```

### Themes 

- You can change the background of the chart 
  + Grey squares is the default
- Other options include:  
  + `+theme_bw()`, `+theme_light()`, `+theme_dark()`, `+theme_classic()`

```{r,}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length,color=Species))+
  geom_point()+
  labs(title="Sepal Length and Petal Length",
      caption="Data from R", 
      x="Sepal Length",y="Petal Length",
      color="Species")+
  theme_bw()
```

### Legends  
- The default legend position is to the right of the plot
- It can be changed using `+theme(legend.position="")`
  + Options are "bottom", "top", "left", or "right"
 - You can also alter the legend name and labels using:  
  + `scale_color_discrete(name="",labels=c("",""))` and
  + `scale_fill_discrete(name="",labels=c("",""))` 
- You can remove one of the legends if you have more than one with `guides(fill="none")` or `guides(color="none")`
  + Need to specify the mapping you want to remove
  - If you want to remove *all* legends you can use `+ theme(legend.position="none")`

  
```{r}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length,color=Species))+
  geom_point()+
  labs(title="Sepal Length and Petal Length",
      caption="Data from R", 
      x="Sepal Length",y="Petal Length",
      color="Species")+
  theme_bw()+
  theme(legend.position="left")

```

## Mini-tasks: Customising your chart {.tabset}
### 1: Colour
- Customise a scatter plot, box plot and density plot with colour
- Scatter plot
  + `ggplot(data=data_frame,aes(x=column_1,y=column_2,color=column_3))+geom_point()`
- Box plot
  + `ggplot(data=ban,aes(x=age,y=F1_lob,color=gender,fill=community))+ geom_boxplot()`
- Density plot
  + `ggplot(data=ban,aes(x=F1_lob,color=age,fill=community))+geom_density()`
  
### 2: Customising with colour
- Customise a scatter plot, box plot and density plot with colour
- Scatter plot
  + `ggplot(data=data_frame,aes(x=column_1,y=column_2,color=column_3))+geom_point()`
- Box plot
  + `ggplot(data=ban,aes(x=age,y=F1_lob,color=gender,fill=community))+ geom_boxplot()`
- Density plot
  + `ggplot(data=ban,aes(x=F1_lob,color=age,fill=community))+geom_density()`

### 3: Labels and Titles 
- Give your plot from Task 4 a new title!
- Try and change the labels of the x and y axes too
  + Previous plot code +`labs(title="Plot Title",caption="Plot Caption",x="column_a",y="column_b", color="column_c",fill="column_d")`
  
### 4: Theme
- Change the theme for the plot from Task 5
  + `+theme_bw()`
  + `+theme_light()`
  + `+theme_dark()`
  + `+theme_classic()`
  
### 5: Legends
- Change the labels for your legend
- Try and remove one of them if you have two

## Faceting {.tabset}
### facet_wrap()
- You can break down your graph further by categorical variables with `facet_wrap()`, which automatically wraps graphs in a rectangle layout
- `facet_wrap(~)` can take one or two arguments
  + for one argument, it goes to the right of the ~ `facet_wrap(~column_e)`
  + for two arguments, one goes on either side of the ~ `facet_wrap(column_e~column_f)`

```{r,eval=TRUE, echo=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length))+
  geom_point()+
  labs(title="Sepal Length and Petal Length",
      caption="Data from R", 
      x="Sepal Length",y="Petal Length",
      color="Species")+
  theme_bw()+
  facet_wrap(~Species)
```

### facet_grid() 
- `facet_grid()` can facet into columns or rows, or both
  + `facet_grid(cols=vars(column_e))` will facet into columns based on this variable
  + `facet_grid(rows=vars(column_f))` will facet into rows based on this variable
  + `facet_grid(cols=vars(column_e), rows=vars(column_f))` will facet into rows and columns based on the two variables

```{r,eval=TRUE, echo=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length))+
  geom_point()+
  labs(title="Sepal Length and Petal Length",
      caption="Data from R", 
      x="Sepal Length",y="Petal Length",
      color="Species")+
  theme_bw()+
  facet_grid(cols=vars(Species))
```

## Changing the colours {.tabset}

- When you map variables onto colour, R will automatically select colours for you
  + But often it's not cute or doesn't match the colour scheme of your presentation
- ggplot lets us change the colours. You can:
  + Manually select each colour within the plot
  + Choose a pre-existing palette 
  
### Manually selecting colours 

- Get a list of the available colour names with `colors()`
- Or you can use the Hex code for the colour
- To change the colours in a plot manually we add `scale_color_manual()`
  + Inside the () we add `values=c()` with the colour names or hex codes in ""
  + When using hex codes, remember to have the # in front of the code

```{r, eval=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length,color=Species))+
  geom_point()+
  scale_color_manual(values=c("blue2","chartreuse2","blueviolet"))
```
  
- Where you have both fill and colour, the same applies  
- To change the fill colour manually add:
  + `+scale_fill_manual(values=c())` 
- To change the line colour manually add:
  + `+scale_color_manual(values=c())`
  
### Pre-existing palettes 

- You don't need to go through choosing colours manually if you don't want to  
- There are a lot of pre-existing palettes that you can add to your graph  
- Different packages have different palettes, I use RColorBrewer   
- We replace the 'manual' in the previous code with `brewer`:  
  + `scale_color_brewer(palette="")`    
  + `scale_fill_brewer(palette="")`  
- You can find palette names in the link in the script  

```{r, eval=TRUE}
ggplot(data=iris, aes(x=Sepal.Length,y=Petal.Length,color=Species))+
  geom_point()+
  scale_color_brewer(palette="RdGy", 
          name="Iris Species")
```

## Mini-tasks: Make it cute and bring it together
### 1: Make it cute
- Manually change the colours in a chart by changing the colours manually, or using a pre-existing palette
_ You can copy and paste a graph from above!

### 2: Bring it together
- Construct a box plot, density plot or scatter plot (i.e. copy and paste from different tasks)
  + Change the theme
  + Change the title and labels (axes and legend)
  + Change the legend (presence or position)
  + Facet it by one or two categorical variables
  + Customise the colours manually or with a pre-existing palette 

  
## Troubleshooting
- Remember to have colour and palette names/hex codes in " " 
- "" and () have to be closed
- R is case-sensitive and space-sensitive 
- there must be a `+` connecting lines 
- Hex codes must have the # before the digits


## Useful Resources

[R Graphics Cookbook](https://r-graphics.org/)  
[R Graph Gallery](https://www.r-graph-gallery.com/ggplot2-package.html)  
[ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/introduction.html)

### Color and palette generators
[Coolers](https://coolors.co/984447-faa6ff-64a6bd-ffba49-1e1e24)
[Colormind](http://colormind.io/)
[Muzli](https://colors.muz.li/)

