---
title: Plotting an s-curve
author: Elena Sheard
date: '2023-04-19'
slug: plotting-an-s-curve
categories: R
tags: visualisation
output:
  blogdown::html_page:
    toc: true
---

Recently I needed to plot an s-curve and it took me an ungodly amount of time to
figure out, because I don't really understand mathematical functions.  

With substantial help from stack overflow I put together this code and adjusted
 it for the sociolinguistics context!

```{r, warning=FALSE}
library(tidyverse)
library(scales)
```
## Basic s-curve

The basic s-curve maps a particular function with `stat_function()` added to a 
ggplot object. I can't explain much more about the function aside from the larger
the numbers in `x` the steeper the line and the more plateau you get at either end of the curve. 

```{r}
#s-curve
p <- ggplot(data = data.frame(x = c(-8, 8)), aes(x))

p +
  stat_function(fun = function(x) exp(x)/(1 + exp(x)), n = 100) +
  theme_bw(base_size = 14)

```

## Adding relevant labels for language change

To relate the curve to language change I used `labs()` to change the x and y axis 
labels to "Year of Birth" (corresponding to older speakers born longer ago and younger speakers born more recently) and "Rate of use of innovative variant". I also changed the
labels of the breaks on the x axis to every 25 years starting from 1900 with
`scale_x_continuous()`. Finally, I just changed the y axis to % using `scale_y_continuous()` and the `scales` package. 

```{r}
p <- ggplot(data = data.frame(x = c(-8, 8)), aes(x))

p +
  stat_function(fun = function(x) exp(x)/(1 + exp(x)), n = 100) +
  labs(x = "Year of Birth",
       y = "Rate of use of innovative variant") +
  scale_x_continuous(breaks = c(-8, -4, 0, 4, 8),
                     labels = c('1900', '1925', '1950',
                                '1975', '2000')) + 
  scale_y_continuous(labels = percent) +
  theme_bw(base_size = 14)
```

